Supplemental Materials for: \\ Large scale risks from pesticides in small streams

### 2017-02-10 article.tex
first draft revisions

\todo{überarbeiten}
% Total words: 199
\end{abstract}

@@

Small streams comprise a major fraction of streams \citep{nadeau_hydrological_2007}, accommodate a higher proportion of biodiversity compared to bigger streams \citep{davies_comparison_2008, biggs_report_2014} and play an important role in the recolonization of disturbed downstream reaches \citep{liess_analyzing_2005, orlinskiy_forested_2015}.
Nevertheless, a clear definition of small streams in terms of catchment or stream size is currently lacking \citep{lorenz_specifics_2016}. 
For example, the WFD defines small streams with a catchment size between 10 and 100 km\textsuperscript{2}, without further categorisation of streams \textless 10km\textsuperscript{2} and \citet{lorenz_specifics_2016} defines small streams with catchment size \textless 10km\textsuperscript{2}. 
Moreover, small streams might particularly be at high risk of pesticide contamination in case of adjacent agricultural areas and given their low dilution potential \citep{schulz_field_2004,liess_determination_1999}.
Indeed, meta-analyses using data from studies with a few sites reported higher pesticide pollution in smaller streams compared to bigger streams \citep{stehle_pesticide_2015,schulz_field_2004}. 
Despite their ecological relevance and potentially higher pesticide exposure, a recent review of pesticide studies showed that a disproportionally small fraction of studies was conducted in small water bodies, and these were largely limited to a few sites \citep{lorenz_specifics_2016}. 
Consequently, knowledge on the pesticide pollution of small streams on larger scales is scant. 
In European law, the Directive 2009/128/EC \citep{European_Union_2009} places an obligation on the EU Member States to adopt National Action Plans (NAP) for the Sustainable Use of Plant Protection Products and the German NAP also addresses the knowledge gap concerning pesticide impact on small streams, specifically including those with catchment size \textless 10km\textsuperscript{2}.

The aim of this study is to identify drivers and dynamics of pesticide concentrations in streams.
To achieve this, we compiled and analysed large-scale pesticide monitoring data from small streams in Germany in order to identify drivers and dynamics of pesticide concentrations. 
We expect that the landscape is a determinant of measured pesticide concentrations.
Because a major fraction of pesticides is applied to agricultural fields, we hypothesised highest concentrations and possible exceedances of RACs in streams with high proportion of agriculture. 
Moreover, if agricultural land use is a main source for pesticides in streams, we expect that exceedances drop to zero if there is not agricultural land use in the catchment. 
Moreover, these relationships may show thresholds that could be to define reference streams without pollution for a future monitoring. 
Given their low dilution potential and direct adjacency to agricultural fields we expected that small streams show highest pesticide concentrations. 
However, also the timing of sampling may influence measured concentrations:
A sampling directly after a precipitation might show higher concentrations because of run-off.
Furthermore, pesticides are not applied throughout the whole year and we expected highest concentrations during the main growing season.
Finally, we quantified the current risks from pesticides in small streams in Germany and the compounds accountable for the risk.

@@

\subsection{Characterization of pesticide pollution}
We characterised pesticide pollution using regulatory acceptable concentrations (RAC) \citep{brock_linking_2010}.
RACs are derived during pesticide authorisation as part of the ecological risk assessment.
No unacceptable ecological effects are expected if the environmental concentration remains below this concentration.
\citet{stehle_pesticide_2015} showed that RAC exceedances reflect a decrease in biodiversity and from this perspective are ecologically relevant indicators. 
The German Environment Agency (UBA) provided RACs for 107 compounds, including those with the highest detection rates (Supplemental Table~S2). 
Based on theses RACs, we calculated Risk Quotients (RQ):

@@

\begin{align}
\begin{split}
\log(\mu_{i}) = log~prec_{0 i} + log~prec_{-1 i} + Q1_{i} + Q2_{i}+Q3_{i}+Q4_{i}\\
logit(\nu_{i}) = log~prec_{0 i} + log~prec_{-1 i} + Q1_{i} + Q2_{i}+Q3_{i}+Q4_{i}
\end{split}
\label{eqn:eqn4}
\end{align}

@@

We fitted this model separately to each compound with a RAC, measured in at least 1000 samples and with more than 5\% of values above LOQ (n = 22 compounds, see Supplemental Table~S3 for a list of compounds). 
To summarise the coefficients across the 22 modelled compounds we used a random effect meta-analysis for each model coefficient separately \citep{harrison_getting_2011}, resulting in an averaged effect of the 22 compounds.
The results of individual compounds are provided in the Supplemental Table~S4 and Figure~S7.
The meta-analysis was performed using the metafor package \citep{Viechtbauer_2010}.

@@

\subsection{Effect of precipitation on pesticide risk}
$prec_{0}$ and $prec_{-1}$ increased the probability of exceeding LOQ and RQ.
In $Q2$ an increase from 0.1~mm to 15~mm of precipitation before sampling ($prec_{-1}$) lead on average to a 43\% higher mean RQ of 0.05 (Supplemental Figure~S7).
The probability to exceed LOQ increases in $Q2$ 1.6-fold from 8.7\% to 13.5\% (Figure~\ref{fig:fig5}). % do_precip.R
Precipitation before sampling ($prec_{-1}$) had a stronger effect than precipitation during sampling ($prec_{0}$) on the probability of exceeding LOQ. 
This difference was less pronounced for the mean value of RQ (Supplemental Figure~S7, topleft). 
Moreover, effects differed between individual compounds (see Supplemental Table~S4).

@@

% current problems in monitoring and possible solutions
A nationwide assessment of pesticide pollution is hampered by inhomogeneous data across federal states:
Beside large differences in the spatial distribution and quantity of sampling sites (Figure~\ref{fig:fig1}), the spectrum of analysed compounds (Figure~\ref{fig:fig2}) and the quality of chemical analyses differed between states. 
Despite the outlined differences between states, all ecoregions occurring in Germany \citep{illies1978limnofauna,abell2008freshwater} were covered by the presented dataset and thus it might nonetheless represent a sample covering all types of small streams in Germany. 
For Thiacloprid and Chlorpyrifos the LOQs were above the RAC, which means that exceedances are likely underestimated.
For compounds with low RACs a lowering of LOQ through an improvement of chemical analysis is essential for reliable assessment.
Moreover, a nationwide assessment would benefit from a harmonised spectrum of analysed compounds between federal states.

@@

\subsection{Influence of agricultural land use and catchment size}
% agriculture
We found a strong influence of agriculture on the pollution of streams.
Above 28\% agriculture within a catchment, it is likely that a RAC will be exceeded, with a further increase in entirely agricultural catchments (above 75 \% agriculture).
To our knowledge, this is the first study investigating such thresholds of pesticide risk.
\todo{more dicussion here?;drop shows the agri is important; compare to linear fit}

@@

\subsection{Effect of precipitation on pesticide risk}
Our results revealed that pesticide sampling for chemical monitoring in Germany is mainly performed when no precipitation occurs. 
Nevertheless, we found a 36\% higher RQ if samples were taken after rainfall events. 
Samples taken on the day of a rainfall event showed less risk than samples taken one day after a rainfall event.
This could be explained by the sampling preceding the rainfall event and the delay between the start of a rain event and the peak in discharge or runoff. 
The effects of precipitation were more pronounced for the probability to exceed LOQ, with smaller effect sizes for the absolute value of RQ. 
This may be explained by a higher variability of absolute concentrations.
Overall, our results indicate that current pesticide monitoring relying on grab sampling, largely disconnected from precipitation events, underestimates pesticide risks.
Automatic event-driven samplers \citep{stehle_probabilistic_2013} and passive samplers \citep{fernandez_calibration_2014,moschet_evaluation_2015} may help overcome these shortcomings and provide a better representation of risks, especially for small water bodies \citep{lorenz_specifics_2016}.

We found the highest the probability of exceeding LOQ during summer (10\% for Q2) and lowest in the first quarter of the year (4\%, Figure~\ref{fig:fig5}, bottom right).
This annual pattern coincides with the main application season for pesticides in Central Europe.
Nevertheless, there are compound-specific differences in the annual pattern, which explains the wide CI for the absolute RQ (Figure~\ref{fig:fig5}, bottom left). 
For example, the herbicide Diflufenican showed the highest RQ and the highest probability of exceeding LOQ during the winter quarters Q1 and Q4 (Supplemental Table~S4), which coincides with the application period it is registered for in Germany \citep{bvl_online_2016}.
Our study suggests that pesticide risks display compound specific spatio-temporal dynamics.
Currently, little is known about these and further research on those might provide useful information for future ecological risk assessment. 
For example, the sensitivity of organisms is often life stage dependent \citep{hutchinson1998analysis} and knowledge on temporal dynamics could inform on concurrent exposure to multiple pesticides, as well as assist to parameterise toxicokinetic and toxicodynamic models \citep{ashauer2016modelling}. 
Moreover, our results show that analysing absolute concentrations and probabilities of LOQ together might deliver valuable insights into risk dynamics.

@@

%%% Word count
% abstract    :  200
% 1 big(600)  :  600
% 5 small(300): 1500
% acknowledg  :   80
%              =====
% subtotal      2380
% text body   : 3846
% ======================================
% subtotal    : 7122
% - captions  :  189
% =========================================
% total       : 6933

>>>>

% Total words: 199
\end{abstract}

@@

Small streams comprise a major fraction of streams \citep{nadeau_hydrological_2007}, accommodate a higher proportion of biodiversity compared to larger streams \citep{davies_comparison_2008, biggs_report_2014} and play an important role in the recolonization of disturbed downstream reaches \citep{liess_analyzing_2005, orlinskiy_forested_2015}.
Nevertheless, a clear definition of small streams in terms of catchment or stream size is currently lacking \citep{lorenz_specifics_2016}. 
For example, the WFD defines small streams with a catchment size between 10 and 100 km\textsuperscript{2}, without further categorisation of streams \textless 10km\textsuperscript{2} and \citet{lorenz_specifics_2016} defines small streams with catchment size \textless 10km\textsuperscript{2}. 
Moreover, small streams might particularly be at high risk of pesticide contamination in case of adjacent agricultural areas and given their low dilution potential \citep{schulz_field_2004,liess_determination_1999}.
Indeed, meta-analyses using data from studies with a few sites reported higher pesticide pollution in smaller streams compared to bigger streams \citep{stehle_pesticide_2015,schulz_field_2004}. 
Despite their ecological relevance and potentially higher pesticide exposure, a recent review of pesticide studies showed that a disproportionally small fraction of studies was conducted in small water bodies, and these were largely limited to a few sites \citep{lorenz_specifics_2016}. 
Consequently, knowledge on the pesticide pollution of small streams on larger scales is scant. 
In European law, the Directive 2009/128/EC \citep{European_Union_2009} places an obligation on the EU Member States to adopt National Action Plans (NAP) for the Sustainable Use of Plant Protection Products and the German NAP also addresses the knowledge gap concerning pesticide impact on small streams, specifically including those with catchment size \textless 10km\textsuperscript{2}.

The aim of this study is to identify drivers and dynamics of pesticide concentrations in streams.
To achieve this, we compiled and analysed large-scale pesticide monitoring data from small streams in Germany in order to identify drivers and dynamics of pesticide concentrations. 
We expect that the landscape is a determinant of measured pesticide concentrations.
Because a major fraction of pesticides is applied to agricultural fields, we hypothesised highest concentrations and possible exceedances of RACs in streams with a high proportion of agriculture. 
Moreover, if agricultural land use is the main source for pesticides in streams, we expect that exceedances drop to zero if there is not agricultural land use in the catchment. 
Moreover, these relationships may show thresholds that could be to define reference streams without pollution for a future monitoring. 
Given their low dilution potential and direct adjacency to agricultural fields we expected that small streams show highest pesticide concentrations. 
However, also the timing of sampling may influence measured concentrations:
A sampling directly after a precipitation might show higher concentrations because of run-off.
Furthermore, pesticides are not applied throughout the whole year and we expected highest concentrations during the main growing season.
Finally, we quantified the current risks from pesticides in small streams in Germany and the compounds accountable for the risk.

@@

\subsection{Characterization of pesticide pollution}
We characterised pesticide pollution using regulatory acceptable concentrations (RAC) \citep{brock_linking_2010}.
RACs are derived during pesticide authorisation as part of the ecological risk assessment (ERA).
According to the goals of ERA, exceedances of RACs should not occur after pesticide authorisation. 
No unacceptable ecological effects are expected if the environmental concentration remains below the RAC. 
\citet{stehle_pesticide_2015} showed that RAC exceedances reflect a decrease in biodiversity and from this perspective are ecologically relevant indicators. 
The German Environment Agency (UBA) provided RACs for 107 compounds, including those with the highest detection rates (Supplemental Table~S2). 
Based on theses RACs, we calculated Risk Quotients (RQ):

@@

\begin{align}
\begin{split}
\log(\mu_{i}) = log(prec_{0 i}) + log(prec_{-1 i}) + Q1_{i} + Q2_{i}+Q3_{i}+Q4_{i}\\
logit(\nu_{i}) = log(prec_{0 i}) + log(prec_{-1 i}) + Q1_{i} + Q2_{i}+Q3_{i}+Q4_{i}
\end{split}
\label{eqn:eqn4}
\end{align}

@@

We fitted this model separately to each compound with a RAC, measured in at least 1000 samples and with more than 5\% of values above LOQ (n = 22 compounds, see Supplemental Table~S3 for a list of compounds). 
To summarise the coefficients across the 22 modelled compounds we used a random effect meta-analysis for each model coefficient separately \citep{harrison_getting_2011}, resulting in an averaged effect of the 22 compounds.
The results of individual compounds are provided in the Supplemental Table~S4 and Figures~S6 and S7.
The meta-analysis was performed using the metafor package \citep{Viechtbauer_2010}.

@@

\subsection{Effect of precipitation on pesticide risk}
$prec_{0}$ and $prec_{-1}$ increased the probability of exceeding LOQ and RQ.
In $Q2$ an increase from 0.1~mm to 15~mm of precipitation before sampling ($prec_{-1}$) lead on average to a 43\% higher mean RQ of 0.05 (Supplemental Figure~S7).
The probability to exceed LOQ increases in $Q2$ 1.6-fold from 8.7\% to 13.5\% (Figure~\ref{fig:fig5}). % do_precip.R
Precipitation before sampling ($prec_{-1}$) had a stronger effect than precipitation during sampling ($prec_{0}$) on the probability of exceeding LOQ. 
This difference was less pronounced for the mean value of RQ (Supplemental Figure~S7, top left). 
Moreover, effects differed between individual compounds (see Supplemental Table~S4).

@@

% current problems in monitoring and possible solutions
A nationwide assessment of pesticide pollution is hampered by inhomogeneous data across federal states:
Beside large differences in the spatial distribution and quantity of sampling sites (Figure~\ref{fig:fig1}), the spectrum of analysed compounds (Figure~\ref{fig:fig2}) and the quality of chemical analyses differed between states. 
Despite the outlined differences between states, all ecoregions occurring in Germany \citep{illies1978limnofauna,abell2008freshwater} were covered by the presented dataset (see Supplemental Figure~S3) and thus it might nonetheless represent a major fraction of small stream types in Germany.

Nevertheless, the differences in monitoring efforts among states hamper a spatial interpolation. 
We accounted for differences in sampling efforts per site, by taking the total number of measurement into account. 
However, we acknowledge that this takes only partly differences between sampling intervals into account which might influence our results \citep{stehle_probabilistic_2013,Xing_Chow_Rees_Meng_Li_Ernst_Benoy_Zha_Hewitt_2013}. 
Moreover, it is known that differences in analytical quality might influence estimated effects \citep{Martin_Eberle_Nakagaki}.
However, the used model (Equation~\ref{eqn:eqn3}) takes explicitly LOQs and differences therein into account.

For Thiacloprid and Chlorpyrifos the LOQs were above the RAC, which means that exceedances are likely underestimated.
For compounds with low RACs a lowering of LOQ through an improvement of chemical analysis is essential for reliable assessment.
Moreover, a nationwide assessment would benefit from a harmonised spectrum of analysed compounds between federal states.

@@

\subsection{Influence of agricultural land use and catchment size}
% agriculture
As hypothesised, we found a positive relationship between agricultural land use and the number of RAC exceedances.
Especially, we found a statistically significant drop below 28\% of agricultural land use (Figure~\ref{fig:fig4}, left).
This drop indicates that agricultural land use might be a major contributor to the observed exceedances.
The absence of such a drop would have indicated that there are exceedances that cannot be explained by agricultural land use, like inputs from urban land uses.

@@

\subsection{Effect of precipitation on pesticide risk}
We found a 36\% higher RQ if samples were taken after rainfall events, that correspond to the hypothesis that pesticides enter streams via run-off.
However, samples taken on the day of a rainfall event showed less risk than samples taken one day after a rainfall event.
This discrepancy could be explained by a sampling preceding the rainfall event because the temporal resolution of our dataset was 1 day.
Additionally, this might be also explained by a delay between the start of a rain event and the peak in discharge or runoff.

The effects of precipitation were more pronounced for the probability to exceed LOQ, with smaller effect sizes for the absolute value of RQ. 
This may be explained by a higher variability of absolute concentrations.
Overall, our results indicate that current pesticide monitoring relying on grab sampling, largely disconnected from precipitation events, underestimates pesticide risks.
Automatic event-driven samplers \citep{stehle_probabilistic_2013} and passive samplers \citep{fernandez_calibration_2014,moschet_evaluation_2015} may help overcome these shortcomings and provide a better representation of risks, especially for small water bodies \citep{lorenz_specifics_2016}.

We found the highest the probability of exceeding LOQ during summer (10\% for Q2) and lowest in the first quarter of the year (4\%, Figure~\ref{fig:fig5}, bottom right).
This annual pattern coincides, as expected, with the main application season for pesticides in Central Europe.
Nevertheless, there are compound-specific differences in the annual pattern, which explains the wide CI for the absolute RQ (Figure~\ref{fig:fig5}, bottom left). 
For example, the herbicide Diflufenican showed the highest RQ and the highest probability of exceeding LOQ during the winter quarters Q1 and Q4 (Supplemental Table~S4), which coincides with the application period it is registered for in Germany \citep{bvl_online_2016}.
Moreover, compound properties, like half-life or water solubility, might influence compound dynamics. 
Our study suggests that pesticide risks display compound specific spatio-temporal dynamics. 
Currently, little is known about these and further research on those might provide useful information for future ecological risk assessment. 
For example, the sensitivity of organisms is often life stage dependent \citep{hutchinson1998analysis} and knowledge on temporal dynamics could inform on concurrent exposure to multiple pesticides, as well as assist to parameterise toxicokinetic and toxicodynamic models \citep{ashauer2016modelling}. 
Moreover, our results show that analysing absolute concentrations and probabilities of LOQ together might deliver valuable insights into risk dynamics.

@@

%%% Word count
% abstract    :  200
% 1 big(600)  :  600
% 5 small(300): 1500
% acknowledg  :   80
%              =====
% subtotal      2380
% text body   : 3912
% ======================================
% subtotal    : 6290
% - captions  :  200
% =========================================
% total       : 6090

### 2017-02-10 supplement/Supplemental_Materials.tex
revised up to discussion
first draft revisions

@@

%% -------------------------------------------------------------------------
\chapter{Thresholds for agricultural land use and catchment size}
\begin{figure}[ht]
	\centering
	\includegraphics[width = 0.8\textwidth]{ezgagrirac}
	\caption[Raw data used for the model in equation 2 and Figure 3 of the main article.]{Raw data used for the model in equation 2 and Figure 3 of the main article. Color codes the number of RAC exceedances (on a log-scale). Grey points denote sites without any exceedance.}
	\label{fig:ezgagrirac}
\end{figure}

@@

%% -------------------------------------------------------------------------
\chapter{Effect of precipitation and season on RQ}
\begin{figure}[ht]
	\centering
	\includegraphics[width = 0.8\textwidth]{precip}
	\caption[Distribution of precipitation at sampling occasions.]{Distribution of precipitation at sampling occasions. top: at sampling date. bottom: at the day before sampling.}
	\label{fig:precip}
\end{figure}

>>>>

\begin{figure}[ht]
	\centering
	\includegraphics[width = 0.8\textwidth]{fig_map}
	\caption{Spatial distribution of the 2,301 small stream sampling sites. Colour codes different federal states (see Supplemental Table S1 for abbreviations). Overlayed are ecoregions. ER4: Alps (no label), ER8: Western highlands,
	ER9: Central highlands, ER 13: Western plains , ER14: Central plains. }
	\label{fig:ecoreg}
\end{figure}

@@

%% -------------------------------------------------------------------------
\chapter{Influence of agricultural land use and catchment size}
\begin{figure}[ht]
	\centering
	\includegraphics[width = 0.8\textwidth]{ezgagrirac}
	\caption[Raw data used for the model in equation 2 and Figure 3 of the main article.]{Raw data used for the model in equation 2 and Figure 3 of the main article. Color codes the number of RAC exceedances (on a log-scale). Grey points denote sites without any exceedance.}
	\label{fig:ezgagrirac}
\end{figure}

@@

%% -------------------------------------------------------------------------
\chapter{Effect of precipitation and season on RQ}
% \begin{figure}[ht]
% 	\centering
% 	\includegraphics[width = 0.8\textwidth]{precip}
% 	\caption[Distribution of precipitation at sampling occasions.]{Distribution of precipitation at sampling occasions. top: at sampling date. bottom: at the day before sampling.}
% 	\label{fig:precip}
% \end{figure}
